---
title: "Cricket Results"
date: \today
# author:  'Apoorva Lal'
# documentclass: amsart
# amsart: true
# thanks: "Acknowledgements: I'd like to thank my mum and dad"
# abstract: '\lipsum[4]'
output:
  html_document:
    theme: flatly
    highlight: tango
    code_folding: hide
    toc: true
    toc_float: true
    toc_depth: 3
    fig_width:  10
    fig_height: 8
    self_contained: true
  # pdf_document:
  #   keep_tex: true
  #   fig_width: 7
  #   fig_height: 6
  #   fig_caption: true
  #   df_print: kable
  #   dev: cairo_pdf
  #   latex_engine: xelatex
  #   template: /home/alal/Templates/boilerplate/lal-mdpdf.latex
# fontsize: 12pt
# colorlinks: true
# indent: true
# xetex: true
# mainfont: TeX Gyre Pagella
# monofont: Iosevka
# mathspec: true
# geometry: margin=2cm
# embed links in citations
# link-citations: true
# numbersections: true
# separate title page
# septitle: true
# spacing
# linestretch: 2
# backend: biber
# bibliography: /home/alal/Dropbox/MyLibrary2.bib
# csl: econometrica.csl
# header-includes:
#   - \usepackage{lipsum}
---

```{r, echo=FALSE, include=FALSE}
# %% ####################################################
rm(list = ls())
library(LalRUtils)
LalRUtils::libreq(tidyverse, data.table, stargazer2,
  lfe, ivmodel, fastDummies, patchwork,
  rio, foreach, magrittr, janitor, knitr, rmarkdown)
theme_set(lal_plot_theme())
options(repr.plot.width=12, repr.plot.height=9)
set.seed(42)
# %% ####################################################
opts_chunk$set(echo=TRUE, autodep=TRUE, cache.comments=FALSE, cache = TRUE,
               message=FALSE, warning=FALSE)
outf = 'html'
```

<!--
+ kappa weights
+ subgroup level hetereogeneous treatment effects
+ wicket to run ratio
 -->

# Outline

```
# Introduction

# Data

# Empirical Specification
+ Instrument estimates the effect of being able to harness the full
  set of information relevant to winning

# Results
## Baseline Treatment Effects
## Heterogeneous Treatment Effects

+ grounds
+ day/night
+ condition on history

## Robustness

# Policy Implications

# Conclusion
```


# Introduction

+ one of the most popular sports in the world
+ unlike other similarly popular sports, it has evolved with new rules
  and formats
  * soccer has remained a 90 minute match in record history, while
    cricket matches can last from anywhere between 3 hours to 5 days
+ common feature of all formats is that there is a team that bats
  first, while the other team fields and bowls
  * in the longer versions of the game, both teams are alloted two
    innings each, while in shorter versions each team bats for an
    innings

 + innings produce an inherent asymmetry of the game because
  * weather conditions may be more conducive to batting
    earlier in the day
  * psychological effects of chasing targets vs being first-mover
+ this asymmetry is mitigated by having a coin toss, in which the team
  that wins the toss gets to choose whether they bat or field first
  * therefore winning the toss can have a profound impact on the odds
    of winning or losing the game that gives one team an advantage in
    win probability against the other team even before a single ball
    is bowled
+ normative question: are these asymmetries created by the toss an
  inherent part of the game that we all accept, or do they detract
  from the integrity and fairness of the game,
  + some people might say that since the toss is random, the
    unfairness is not objectionable, since teams eventually
  + however, some matches (such as knock-out matches in world cups)
    are of such consequences that their outcome ought to depend on the
    skill, stamina, training, decisionmaking, and coaching and
    leadership of captain etc rather than the outcome of a random
    event that occurs before a single ball is bowled, before a single
    ball is bowled, so the argument goes

+ In this paper, we investigate the impacts of these asymmetries on
  match outcomes  using an original dataset of 60,000 cricket matches
  played over the course of the last century. This is the most
  comprehensive cricket match dataset to our knowledge

+ Our analysis proceeds in the following series of steps
  * is the toss random?
  * what is the effect of winning the toss on winning the game?
    - 2%
    - team strength
    - is it higher for more competitive games?
    - match types
    - how does this vary by the competing teams' characteristics
    - stakes of the tournament
  * since the toss endows one of the captains with the choice of
    batting first or second, how do captains use the toss to their
    advantage
  * in other words, what is the effect of having the choice to bat
    first and choosing to do so?
    - hetergeneous effects
    - IV
    - Kappa
+ can we do this in terms of runs? this is the policy relevant
  quantity


# Data

## Data Sanity Checks

```{r}
# %% long data
root = '/home/alal/Dropbox/1_Research/cricket/'
datadir = file.path(root, "repos/cricket-stats/data")
crickett = fread(file.path(root, "tmp/matches_long.csv"))
setorder(crickett, -year, -month, -date, mid)
```

```{r, echo=F, include = F}
# check that match id is unique
crickett[, .N, mid][order(-N)][1]
# dedupe
# %% # ## Subsample (for now)
crickett$type_of_match %>% tabyl
regsamp = crickett[type_of_match  %in%
  c("ODI", "T20", "T20I", "Twenty20", "TEST", "FC", "LISTA", "Women's T20")]
# %% throw out matches with no results / cancelled / walkover
regsamp[outcome %like% 'abandoned' | outcome %like% 'cancelled', .N]
regsamp = regsamp[!(outcome %like% 'abandoned' | outcome %like% 'cancelled')]
regsamp[outcome %like% 'walkover' , .N]
regsamp = regsamp[!(outcome %like% 'walkover')]
# fix bat first and toss indicators
########     ###    ########  #######  ########  ########  ######## ########
##     ##   ## ##      ##    ##     ## ##     ## ##     ## ##       ##     ##
##     ##  ##   ##     ##    ##     ## ##     ## ##     ## ##       ##     ##
########  ##     ##    ##    ##     ## ########  ##     ## ######   ########
##     ## #########    ##    ##     ## ##   ##   ##     ## ##       ##   ##
##     ## ##     ##    ##    ##     ## ##    ##  ##     ## ##       ##    ##
########  ##     ##    ##     #######  ##     ## ########  ######## ##     ##
regsamp$bat_first %>% tabyl
# win toss checks
regsamp$wintoss %>% tabyl
# %% final batting order checks
regsamp[, bat_first_min := min(bat_first), by = mid]
regsamp$bat_first_min %>% tabyl
# these have no toss info
regsamp[bat_first_min == 1,
  .(mid, outcome, wintoss, wingame, bat_first, bat_first_min)]
# drop them
regsamp = regsamp[!(bat_first_min == 1)]

# %% check coding of draws
regsamp[, .N, wingame]

# %%
# some matches record ties differently
regsamp[, res_match_exists := max(wingame), mid] # matches for which at least one team is recorded to have won
regsamp[, table(res_match_exists)]
# draws for test and FC matches
regsamp[res_match_exists == 0.5, table(type_of_match)]
# %%
regsamp[res_match_exists == 0, .N, type_of_match]
regsamp[res_match_exists == 0, .(outcome)] %>% head
# %% recode them to draw for now - check w others later
regsamp[res_match_exists == 0, wingame := 0.5]

# %% finally, consolidate match type before FEs and subgroup analyses
regsamp[, .N, type_of_match]

regsamp[, type_of_match2 := case_when(
  type_of_match == "ODI"          ~ "OD",
  type_of_match == "T20I"         ~ "T20",
  type_of_match == "Women\'s T20" ~ "T20",
  TRUE ~ type_of_match)]

regsamp %>% tabyl(type_of_match2)


# %% time indicators
regsamp[, time_unit  := cut(year,
  breaks = c(-Inf, 1850, 1900, 1950, 1980, 1990, 2000, 2010, 2020),
  labels = c("<=1850", "1851-1900", "1901-1950", "1951-1980" ,"1981-1990",
             "1991-2000", "2001-2010", "2011-2020"))]
# regsamp[, tabyl(year), by = time_unit]

# %% sanity checks - these should be perfectly balanced
regsamp[, .N, bat_first]
regsamp[, .N, wingame]
regsamp[, .N, wintoss]

```

# Empirics

## Pooled Estimates

```{r}
# %% overall
ols          = felm(wingame ~ bat_first | 0 | 0 | mid , regsamp)
# matches where toss-winner chose to bat first
ols0         = felm(wingame ~ bat_first | 0 | 0 | mid , regsamp[bat_or_bowl == 1])
# matches where toss-winner chose to field first
ols1         = felm(wingame ~ bat_first | 0 | 0 | mid , regsamp[bat_or_bowl == 2])
first_stage  = felm(bat_first ~ wintoss | 0 | 0 | mid , regsamp)
reduced_form = felm(wingame ~ wintoss | 0 | 0 | mid , regsamp)
iv_est       = felm(wingame ~ 1 | 0 | (bat_first ~ wintoss) | mid , regsamp)
iv_est_2     = felm(wingame ~ 1 |
                    as.factor(name) + as.factor(type_of_match2) | (bat_first ~ wintoss) | mid, regsamp)
iv_est_3     = felm(wingame ~ 1 |
                    as.factor(name) + as.factor(type_of_match2) + as.factor(time_unit) |
                    (bat_first ~ wintoss) | mid, regsamp)
# %%
```


\bland
```{r, results='asis'}
reglabs = c("Bat First", "Win Toss", "Bat First (IV)")
marker = function (name, yesno){
        return(c(name, map_chr(yesno, function(x) ifelse(x, "✓", ""))))
}

fstatrow = function(m) round(m$stage1$iv1fstat[[1]][['F']], 2)

stargazer(ols, ols0, ols1, first_stage, reduced_form,
          iv_est, iv_est_2, iv_est_3,
  add.lines = list(c("Sample",
                    "All", "TW bats", "TW fields", "All", "All", "All", "All", "All"),
                  c("FS F-Stat", "", "", "", "", "",
                      fstatrow(iv_est), fstatrow(iv_est_2), fstatrow(iv_est_3)),
                  marker('Team FE',       c(F, F, F, F, F, F, T, T)),
                  marker('Match-Type FE', c(F, F, F, F, F, F, T, T)),
                  marker('Decade FE',     c(F, F, F, F, F, F, F, T))) ,
  keep.stat = c('N'),
  covariate.labels = reglabs,
  dep.var.labels.include = FALSE,
  column.labels   = c("OLS", "RF", "FS", "IV"),
    column.separate = c(3, 1, 1, 3),
  type = outf)
```
\eland

```{r}
reg_by_group = function(df, groupvar, fml){
  require(dplyr) ; require(purrr) ; require(broom)
  group_reg_tables = df %>%
    group_by({{groupvar}}) %>%
    group_map( ~ robustify(felm(fml, .x)) %>%
      tidy %>% mutate(group = .y[[1]]), .keep = T)
  return(group_reg_tables)
}

coefplotter = function(df, title){
  df %>%
    mutate(term2 = paste0(group, '_', term)) %>%
    ggplot(. , aes(y = group, x = estimate) ) +
      geom_point(colour = '#6495ed') +
      geom_pointrange(aes(
        xmin = estimate - 1.96 * `std.error`,
        xmax = estimate + 1.96 * `std.error`,
      ), colour = "#6495ed", alpha = 0.8) +
    labs(title = title, x = "", y = "") +
    geom_vline(xintercept = 0, linetype = 'dotted')
}
```

## Subgroup Analysis

### Reduced form Effects

```{r}
reg_by_group(regsamp, type_of_match2,
    formula_lfe('wingame', 'wintoss', C = "mid")) %>% rbindlist %>%
  .[term == 'wintoss'] %>%
  coefplotter(., "Reduced Form : Effect of Toss on Win Probability")

reg_by_group(regsamp, di_type_of_match,
    formula_lfe('wingame', 'wintoss', C = "mid")) %>% rbindlist %>%
  .[term == "wintoss"] %>%
  coefplotter(., "Reduced Form - by Intl Category")

reg_by_group(regsamp, time_unit,
    formula_lfe('wingame', 'wintoss', C = "mid")) %>% rbindlist %>%
  .[term == "wintoss"] %>%
  coefplotter(., "Reduced Form - over time") + coord_flip()


reg_by_group(regsamp, continent,
    formula_lfe('wingame', 'wintoss', C = "mid")) %>% rbindlist %>%
  .[term == "wintoss"] %>%
  coefplotter(., "Reduced Form - by continent")

regsamp[, dn := 1*(day_n_night == "day/night match")]

regsamp[type_of_match == "ODI"]  %>%
  reg_by_group(., dn,
    formula_lfe('wingame', 'wintoss', C = "mid")) %>% rbindlist %>%
  .[term == "wintoss"] %>%
  coefplotter(., "Reduced Form - by Day-Night")

```

### First stage

+ History of team + ranking
+ Micro-seasonality
+ Team-FEs

```{r}
reg_by_group(regsamp, type_of_match2,
    formula_lfe('bat_first', 'wintoss', C = "mid")) %>% rbindlist %>%
  .[term == "wintoss"] %>%
  coefplotter(., "First Stage - Effect of Toss on Choice to bat first")

reg_by_group(regsamp, di_type_of_match,
    formula_lfe('bat_first', 'wintoss', C = "mid")) %>% rbindlist %>%
  .[term == "wintoss"] %>%
  coefplotter(., "First Stage - by intl type")

reg_by_group(regsamp, time_unit,
    formula_lfe('bat_first', 'wintoss', C = "mid")) %>% rbindlist %>%
  .[term == "wintoss"] %>%
  coefplotter(., "First Stage - over time") + coord_flip()

reg_by_group(regsamp, continent,
    formula_lfe('bat_first', 'wintoss', C = "mid")) %>% rbindlist %>%
  .[term == "wintoss"] %>%
  coefplotter(., "First Stage - by Continent")
```

### IV by match type


+ History of team

```{r}
# est = felm(wingame ~ 1 | 0 | (bat_first ~ wintoss) | mid , regsamp[type_of_match2 == "LISTA"])
# est
reg_by_group(regsamp, type_of_match2,
      as.formula("wingame ~ 1 | 0 | (bat_first ~ wintoss)| mid")) %>% rbindlist %>%
  .[term == "`bat_first(fit)`"] %>%
  coefplotter(., "IV Estimate - Effect of Batting First on Win Probability")

reg_by_group(regsamp, di_type_of_match,
    as.formula('wingame ~ 1 | 0 | (bat_first ~ wintoss) | mid')) %>% rbindlist %>%
  .[term == "`bat_first(fit)`"] %>%
  coefplotter(., "IV Estimate - by Intl Category")

reg_by_group(regsamp, time_unit,
    as.formula('wingame ~ 1 | 0 | (bat_first ~ wintoss) | mid')) %>% rbindlist %>%
  .[term == "`bat_first(fit)`"] %>%
  coefplotter(., "IV Estimate - over Time") + coord_flip()


reg_by_group(regsamp, continent,
    as.formula('wingame ~ 1 | 0 | (bat_first ~ wintoss) | mid')) %>% rbindlist %>%
  .[term == "`bat_first(fit)`"] %>%
  coefplotter(., "IV Estimate - by Continent")

```

## Characterising Compliers

$$
P(D_1 > D_0) = E[D_1 - D_0] = E[D | Z = 1] - E[D | Z = 0]
$$

```{r}
mu = regsamp[, mean(bat_first), wintoss][, V1]
always_takers = nrow(regsamp[wintoss != 1 & bat_first == 1]) / nrow(regsamp[wintoss != 1])
never_takers =  nrow(regsamp[wintoss == 1 & bat_first == 0]) / nrow(regsamp[wintoss == 1])
compliers = mu[2] - mu[1]


shares = data.frame(subpop = c('always takers', 'never takers', 'compliers'),
           share  = c(always_takers, never_takers, compliers[1]))
shares %>% kable()
```


## $\kappa$ weighted summary statistics


The relative likelihood that a complier has $X = x$ is given by the ratio of the
first stage for the subpopulation with $X = x$ to the overall first stage.

Target: mean for compliers

$$
E[X | D_1 > D_0] = \frac{E[\kappa X]}{E[\kappa]}
$$

where

$$
\kappa = 1 - \frac{D (1-Z)}{P(Z = 0 | X)} - \frac{(1-D) Z}{P(Z = 1 | X)}
$$

+ $\approx$ 1 for identifiable compliers
+ $\approx$ -1 for identifiable always takers
+ $\approx$ -1 for identifiable never takers

Covars considered:

+ underdog - lower ranked team in match
+ day-night matches
+ home team

```{r}
samp2 = regsamp[!is.na(rank) & !is.na(home_country)]
samp2[, lowrank := min(rank) , by = mid]
samp2[, underdog := ifelse(rank == lowrank, 1, 0)]
samp2[, dn := 1*(day_n_night == "day/night match")]
samp2[, home := 1*home_country]
samp2 = dummy_cols(samp2, select_columns = c('time_unit', 'continent')) %>%
  clean_names()


kappa_avg = function(xn, df = samp2) {
  x = as.matrix(df[, ..xn]); y = as.matrix(df$wingame)
  d = df$bat_first; z = df$wintoss
  ϕ = glm(z ~ x, family = binomial())$fitted.values
  κ = 1 - (d * (1-z))/(1-ϕ) - ((1-d) * z)/ϕ
  mean(κ * x) / mean(κ)
}

samp2 %>% glimpse


glue::glue("underdog share among compliers = {round(kappa_avg('underdog'), 3)}")
glue::glue("day-night share among compliers = {round(kappa_avg('dn'), 3)}")
glue::glue("home share among compliers = {round(kappa_avg('home'), 3)}")
glue::glue("DL among compliers = {round(kappa_avg('duckworth_lewis'), 3)}")

times = colnames(samp2) %>% str_subset("time_unit_")
time_kappa = times %>% map_dbl(kappa_avg)
cbind(times, time_kappa)

```


## Local Average Response Function

CEF of $Y|X,D$ *for subpopulation of compliers*.

Target : $E[Y | X, D, D_1 > D_0]$.

Estimate using Kappa weighting.

```{r}

larf_fit = function(xn, df = samp2) {
  X = as.matrix(df[, ..xn]); y = as.matrix(df$wingame)
  d = df$bat_first; z = df$wintoss
  ϕ = glm(z ~ X, family = binomial())$fitted.values
  κ = 1 - (d * (1-z))/(1-ϕ) - ((1-d) * z)/ϕ
  Xw = cbind(1, X, d)
  larf <- solve(t(Xw) %*% diag(κ) %*% Xw ) %*% t(Xw) %*% diag(κ) %*% y
  return(larf)
}

larf_fit(c('underdog', 'dn', 'home'))
```

# Run Curves

```{r}
library(patchwork)
df  = import(file.path(root, "tmp/overs_odi/overs/overs_1000891.json")) %>% setDT
df1 = import(file.path(root, "tmp/overs_odi/overs/overs_1000893.json")) %>% setDT
# %%
cum_run_curves = function(df){
  subt = paste(unique(df$teamShortName) , collapse = ' vs ')
  df[, remaining_wickets := 11 - totalWicket]
  # max run tallies
  totruns = df[, max(totalRuns), by = 'innings']
  # figure
  f = ggplot(df, aes(over, totalRuns, colour = as.factor(innings))) +
    geom_point(aes(alpha = remaining_wickets)) +
    geom_hline(yintercept = totruns[1, 2][[1]], colour = 'red', alpha = 0.5, linetype = 'dashed') +
    geom_hline(yintercept = totruns[2, 2][[1]], colour = 'blue' , alpha = 0.5, linetype = 'dashed') +
    geom_vline(xintercept = 50, alpha = 0.5, linetype = 'dotted') +
    scale_colour_brewer(palette = "Set1") +
    labs(subtitle = subt, caption = 'alpha scaled by number of wickets left',
      colour = 'Innings') + scale_alpha(guide = 'none')
  return(f)
}

p1 = cum_run_curves(df1) + ggtitle('Failed Chase')
p2 = cum_run_curves(df) + ggtitle('Successful Chase')

# %%
p = p1 / p2
p
```

```{r}
matchdata = fread(file.path(root, 'repos/cricket-stats/data/final_output.csv'))
ODIs = matchdata[type_of_match == "ODI"]
dl_matches = ODIs[duckworth_lewis == 1]
dl_ids = dl_matches[, match_id]
dl_id = dl_ids[[50]]
dl_matches[match_id == dl_id, outcome]

# %%
match = import(file.path(root, glue::glue("tmp/overs_odi/overs/overs_{dl_id}.json"))) %>% setDT

match %>% glimpse

cum_run_curves(match)
```
